The Magnetorotational Instability Prefers Three Dimensions

### 2019-08-25 paper/paper.tex
better organise the equations a bit.

We numerically solve the linearized magnetohydrodynamic equations in rotating plane-Couette geometry.
This corresponds to a Cartesian frame rotating with angular frequency $\Omega$ and a linear background shear, $V_{y}=Sx$; \citep[see][]{2015RSPSA.47140699V}.
We cast the Navier-Stokes equation in the form,
\begin{equation}\label{eq:mhd}
\frac{D \boldsymbol{v}}{Dt}+f \boldsymbol{\hat{z}}\times\boldsymbol{v}+{S}v_{x}\,\boldsymbol{\hat{y}}+\boldsymbol{\nabla}{p}+\nu\,\boldsymbol{\nabla}\times\boldsymbol{\omega}=B_{0}\partial_{z}\boldsymbol{b},
\end{equation}
where
\begin{equation}
\boldsymbol{\omega}=\boldsymbol{\nabla}\times\boldsymbol{v},\quad\text{and}\quad\frac{D}{Dt}=\partial_{t}+{S}x\,\partial_{y}\end{equation}
The main input parameters are the Coriolis parameter, $f=2 \Omega$; the background shear rate, $S=dV_{y}/dx<0$;  the vertical magnetic field $B_{0}$ (in Alfv\'{e}n units $\mu_{0}\rho_{0}=1$); and the horizontal domain width $-d/2\le{x}\le d/2$.
Accretion-disk modelling usually considers the Rossby number $q=-S/f<1$; $q=1$ corresponds to purely hydrodynamical Rayleigh unstable shear.
\textit{Note}: some sources use $q=-S/\Omega=3/2$ to represent the same quantity. 
Unless otherwise stated $q=3/4$ (Keplerian).
We write the induction equation in terms of the $x$-component magnetic field,
\begin{equation}\label{eq:Bx}
\frac{Db_{x}}{Dt}+\eta(\partial_{y}j_{z}-\partial_{z}j_{y})=B_{0}\partial_{z}v_{x},
\end{equation}
and the $x$-component current density ($j_{x}=\partial_{y}b_{z}-\partial_{z}b_{y}$),
\begin{equation}\label{eq:Jx}
\frac{Dj_{x}}{Dt}-\eta\nabla^{2}j_{x}=B_{0}\partial_{z}\omega_{x}-S\partial_{z}b_{x}.
\end{equation}
We explicitly enforce divergence-free velocity and magnetic field,
\begin{equation}\label{eq:divu}
 \boldsymbol{\nabla}\cdot\boldsymbol{v}=\boldsymbol{\nabla}\cdot\boldsymbol{b}=0.
\end{equation}
The spatial domain is a doubly periodic channel in $y,\,z$ with width $d=\pi$. 
The boundary conditions are impenetrable stress-free and perfectly conducting; $v_{x}=\omega_{y}=\omega_{z}=b_{x}=\partial_{x}j_{x}=0$ at $x=\pm{d}$. 
The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in \textit{two dimensions} is
\begin{equation}\label{eq:Sc}
  S_{c}=-\frac{\pi^{2}B_{0}^2}{fd^2}.
\end{equation}
We use $\SSC$ as our instability control parameter.

@@

Our two main results are: 
(I) When the first two-dimensional mode becomes unstable, there already exist three-dimensional modes with positive growth rate. 
(II) At sufficiently large criticality, the fastest-growing mode becomes purely two-dimensional. 
These results are pertinent in two ways. 
First, the MRI contains a substantial ``Goldilocks regime" with possible direct dynamo action, and this regime likely applies to stellar interiors and laboratory experiments. 
Second, our results accord with well-established results for accretion disks that expect two-dimensional primary modes.
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC=1.02$, the maximum growth rate ($\gamma/|S|=0.093$) happens at $(k_{y},k_{z})\simeq(0.263,0.447)$.
The first panel of figure~\ref{fig:growth_rate}a shows $\SSC=0.640$, which contains only three-dimensional instability.
The overall critical shear for three-dimensional modes is $S_{c,\text{3D}}\simeq0.102S_c$.

@@

Figure~\ref{fig:eigvec} shows the eigenvectors for the most unstable mode at $\SSC=1.02$ using ideal MHD. 
No critical layers can form for $\gamma\ne0$.
We therefore solve equations~(\ref{eq:mhd}--\ref{eq:divu}) without dissipation as a 2nd-order system in $\partial_{x}$ and impose $v_{x}=0$ at $x=\pm{d}$.
The eigenfunctions are indistinguishable from those at finite dissipation, lending additional confidence to our other results.
The tilted structures of $v_{x}$ and $v_{y}$ as well as $b_{x}$ and $b_{y}$, imply non-trivial Reynolds and Maxwell stresses.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\columnwidth]{run_47_output_growthrates.pdf}\\
  \includegraphics[width=\columnwidth]{run_49_output_growthrates.pdf}
  \caption{(top) Growth rates for $q=1/10$ relevant to the stellar interiors. 
  (bottom) Growth rates for MRI near onset at liquid metal-like diffusivities, $\nu=10^{-6}$, $\eta=0.1$.}
  \label{fig:other_params}
\end{figure}
Finally, we vary two important parameters:
$q$ and the magnetic Prandtl number $\Prm=\nu/\eta$.
Figure~\ref{fig:other_params} shows the growth rates for $\SSC=1.02$.
The upper panel shows $q=1/10$ (addressing a possible range within stellar interiors) holding all other parameters equal to their fiducial values.
The growth rates show similar behavior to the $q=3/4$ cases, \textit{i.e.}, dominated by three-dimensional modes with $\gamma_{\text{max}}/|S|\simeq0.1$.
The lower panel shows $\nu=10^{-6}$ and $\eta=0.1$ with all other parameters equal to their fiducial values. 
This case is relevant to liquid metals e.g., the Princeton MRI experiment \citep{2002JFM...462..365G}
at $\Prm=10^{-5}$. 
Our results are for the rotating plane-Couette geometry (Taylor-Couette in the small-gap limit $R_{2}\approx{R}_{1}$) with highly idealized boundary conditions.
It is nevertheless quite interesting that, near onset, the low-$\Prm$ MRI is \emph{only} unstable to three-dimensional modes.

>>>>

We numerically solve the linearized magnetohydrodynamic equations in rotating plane-Couette geometry.
This corresponds to a Cartesian frame rotating with angular frequency $\Omega$ and a linear background shear, $V_{y}=Sx$; \citep[see][]{2015RSPSA.47140699V}.
We cast the Navier-Stokes equation in the form,
\begin{equation}\label{eq:mhd}
\frac{D \boldsymbol{v}}{Dt}+f \boldsymbol{\hat{z}}\times\boldsymbol{v}+{S}v_{x}\,\boldsymbol{\hat{y}}+\boldsymbol{\nabla}{p}+\nu\,\boldsymbol{\nabla}\times\boldsymbol{\omega}=B_{0}\partial_{z}\boldsymbol{b},
\end{equation}
where
\begin{equation}
\boldsymbol{\omega}=\boldsymbol{\nabla}\times\boldsymbol{v},\quad\text{and}\quad\frac{D}{Dt}=\partial_{t}+{S}x\,\partial_{y}\end{equation}

We write the induction equation in terms of the $x$-component magnetic field,
\begin{equation}\label{eq:Bx}
\frac{Db_{x}}{Dt}+\eta(\partial_{y}j_{z}-\partial_{z}j_{y})=B_{0}\partial_{z}v_{x},
\end{equation}
and the $x$-component current density ($j_{x}=\partial_{y}b_{z}-\partial_{z}b_{y}$),
\begin{equation}\label{eq:Jx}
\frac{Dj_{x}}{Dt}-\eta\nabla^{2}j_{x}=B_{0}\partial_{z}\omega_{x}-S\partial_{z}b_{x}.
\end{equation}
We explicitly enforce divergence-free velocity and magnetic field,
\begin{equation}\label{eq:divu}
 \boldsymbol{\nabla}\cdot\boldsymbol{v}=\boldsymbol{\nabla}\cdot\boldsymbol{b}=0.
\end{equation}
The spatial domain is a doubly periodic channel in $y,\,z$ with width $-d/2\le{x}\le d/2$.
The boundary conditions are impenetrable stress-free and perfectly conducting; $v_{x}=\omega_{y}=\omega_{z}=b_{x}=\partial_{x}j_{x}=0$ at $x=\pm{d/2}$.

The main input parameters are the Coriolis parameter, $f=2 \Omega$; the background shear rate, $S=dV_{y}/dx<0$;  the vertical magnetic field $B_{0}$ (in Alfv\'{e}n units $\mu_{0}\rho_{0}=1$).
Accretion-disk modelling usually considers the Rossby number $q=-S/f<1$; $q=1$ corresponds to purely hydrodynamical Rayleigh unstable shear.
\textit{Note}: some sources use $q=-S/\Omega$ to represent same effective quantity. 
Unless otherwise stated $q=3/4$ (Keplerian).

The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in \textit{two dimensions} is
\begin{equation}\label{eq:Sc}
  S_{c}=-\frac{\pi^{2}B_{0}^2}{fd^2}.
\end{equation}
We use $\SSC$ as our instability control parameter.

@@

Our two main results are: 
(I) When the first two-dimensional mode becomes unstable, there already exist three-dimensional modes with positive growth rate. 
(II) At sufficiently large criticality, the fastest-growing mode becomes purely two-dimensional. 
These results are pertinent in two ways. 
First, the MRI contains a substantial ``Goldilocks regime" with possible direct dynamo action, and this regime likely applies to stellar interiors and laboratory experiments. 
Second, our results accord with well-established results for accretion disks that expect two-dimensional primary modes.
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC=1.02$, the maximum growth rate ($\gamma_{\max}/|S|=0.093$) happens at $(k_{y},k_{z})\simeq(0.263,0.447)$.
The first panel of figure~\ref{fig:growth_rate}a shows $\SSC=0.640$, which contains only three-dimensional instability.
The overall critical shear for three-dimensional modes is $S_{c,\text{3D}}\simeq0.102S_c$.

@@

Figure~\ref{fig:eigvec} shows the eigenvectors for the most unstable mode at $\SSC=1.02$ using ideal MHD. 
No critical layers can form for $\gamma\ne0$.
We therefore solve equations~(\ref{eq:mhd}--\ref{eq:divu}) without dissipation as a 2nd-order system in $\partial_{x}$ and impose $v_{x}=0$ at $x=\pm{d/2}$.
The eigenfunctions are indistinguishable from those at finite dissipation, lending additional confidence to our other results.
The tilted structures of $v_{x}$ and $v_{y}$ as well as $b_{x}$ and $b_{y}$, imply non-trivial Reynolds and Maxwell stresses.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\columnwidth]{run_47_output_growthrates.pdf}\\
  \includegraphics[width=\columnwidth]{run_49_output_growthrates.pdf}
  \caption{(top) Growth rates for $q=1/10$ relevant to the stellar interiors. 
  (bottom) Growth rates for MRI near onset at liquid metal-like diffusivities, $\nu=10^{-6}$, $\eta=0.1$.}
  \label{fig:other_params}
\end{figure}
Finally, we vary two important parameters:
$q$ and the magnetic Prandtl number $\Prm=\nu/\eta$.
Figure~\ref{fig:other_params} shows the growth rates for $\SSC=1.02$.
The upper panel shows $q=1/10$ (addressing a possible range within stellar interiors) holding all other parameters equal to their fiducial values.
The growth rates show similar behavior to the $q=3/4$ cases, \textit{i.e.}, dominated by three-dimensional modes with $\gamma_{\max}/|S|\simeq0.1$.
The lower panel shows $\nu=10^{-6}$ and $\eta=0.1$ with all other parameters equal to their fiducial values. 
This case is relevant to liquid metals e.g., the Princeton MRI experiment \citep{2002JFM...462..365G}
at $\Prm=10^{-5}$. 
Our results are for the rotating plane-Couette geometry (Taylor-Couette in the small-gap limit $R_{2}\approx{R}_{1}$) with highly idealized boundary conditions.
It is nevertheless quite interesting that, near onset, the low-$\Prm$ MRI is \emph{only} unstable to three-dimensional modes.
